Shift-sensitive Power Transformations for Transforming Data to Normality

### 2023-02-01 manuscript/manuscript.Rmd
Update manuscript.Rmd

```{r reduced-normality, echo=FALSE, fig.cap=cap}
cap <- paste0(
  "Power transformations may reduce normality depending on location. ",
  "Variable $x$ consists of $10000$ values are randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$. ",
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly dependent on the location $\\mu$, ",
  "and even breaks down due to numerical issues for $\\mu > 10^3$. For values $\\mu \\leq 10^3$, $\\lambda$ values are similar between methods.")

@@

x <- stats::rnorm(10000L)
shift_range <- 10^seq(from=0, to=6, by=0.1)

@@

p_bc <- ggplot2::ggplot(
  data = data[method == "Box-Cox"],
  mapping = ggplot2::aes(
    x = mu,
    y = lambda,
    colour = method))
p_bc <- p_bc + plot_theme
p_bc <- p_bc + ggplot2::geom_point()
p_bc <- p_bc + ggplot2::geom_hline(
  yintercept = 1.0, 
  linetype="longdash", 
  colour="grey40")
p_bc <- p_bc + ggplot2::annotate(
  geom = "text",
  x = 6,
  y = 1,
  label = "expected",
  colour=annotation_settings$colour,
  family=annotation_settings$family,
  fontface=annotation_settings$face,
  size=annotation_settings$geom_text_size,
  vjust=-1.0,
  hjust=1.0)
p_bc <- p_bc + ggplot2::scale_x_continuous(
  name = latex2exp::TeX("$\\mu$"),
  labels = scales::math_format())
p_bc <- p_bc + ggplot2::scale_y_continuous(
  name = latex2exp::TeX("$\\lambda$"),
  limits = c(-3.6, 1.1))
p_bc <- p_bc + paletteer::scale_color_paletteer_d(
  palette="ggthemes::Tableau_10",
  drop=FALSE,
  guide="none")

p_yj <- ggplot2::ggplot(
  data = data[method == "Yeo-Johnson"],
  mapping = ggplot2::aes(
    x = mu,
    y = lambda,
    colour = method))
p_yj <- p_yj + plot_theme
p_yj <- p_yj + ggplot2::geom_point()
p_yj <- p_yj + ggplot2::geom_hline(
  yintercept = 1.0,
  linetype="longdash",
  colour="grey40")
p_yj <- p_yj + ggplot2::annotate(
  geom = "text",
  x = 6,
  y = 1,
  label = "expected",
  colour=annotation_settings$colour,
  family=annotation_settings$family,
  fontface=annotation_settings$face,
  size=annotation_settings$geom_text_size,
  vjust=-1.0,
  hjust=1.0)
p_yj <- p_yj + ggplot2::scale_x_continuous(
  name = latex2exp::TeX("$\\mu$"),
  labels = scales::math_format())
p_yj <- p_yj + ggplot2::scale_y_continuous(
  name = latex2exp::TeX("$\\lambda$"),
  limits = c(-3.6, 1.1))
p_yj <- p_yj + paletteer::scale_color_paletteer_d(
  palette="ggthemes::Tableau_10",
  drop=FALSE)
p_yj <- p_yj + ggplot2::theme(
  axis.title.y=ggplot2::element_blank(),
  axis.text.y=ggplot2::element_blank())

@@

@@

```{r shifted-distributions, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=cap, include=FALSE, eval = FALSE}
cap <- paste0(
  "Power transformations modified to be shift-sensitive produce transformation parameters are invariant to location.")

@@

        for(method in c("box_cox", "yeo_johnson")){
          for(shift in c(FALSE, TRUE)){
            for(robust in c(FALSE, TRUE)){
              # Skip if robust, but not shifted.
              if(robust && !shift) next

@@

# Update distribution, method and version to factors.
data$distribution <- factor(
  x = data$distribution,
  levels = c("normal", "right-skewed", "left-skewed"))
data$method <- factor(
  x = data$method,
  levels = c("box_cox", "yeo_johnson"),
  labels = c("Box-Cox", "Yeo-Johnson"))
data$version <- factor(
  x = data$version,
  levels = c("original", "shift-sensitive", "robust shift-sensitive"))

@@

# Box-Cox
p_bc_normal <- .create_lambda_shift_plot(
  data = data[distribution == "normal" & method == "Box-Cox"],
  plot_theme = plot_theme,
  limits = c(-3.6, 1.1))

# Yeo-Johnson
p_yj_normal <- .create_lambda_shift_plot(
  data=data[distribution == "normal" & method == "Yeo-Johnson"],
  plot_theme = plot_theme,
  limits = c(-3.6, 1.1))

@@

# Box-Cox
p_bc_right <- .create_lambda_shift_plot(
  data =data[distribution == "right-skewed" & method == "Box-Cox"],
  plot_theme = plot_theme,
  limits = c(1.0, 65))

# Yeo-Johnson
p_yj_right <- .create_lambda_shift_plot(
  data = data[distribution == "right-skewed" & method == "Yeo-Johnson"],
  plot_theme = plot_theme,
  limits = c(1.0, 65))

@@

# Box-Cox
p_bc_left <- .create_lambda_shift_plot(
  data = data[distribution == "left-skewed" & method == "Box-Cox"],
  plot_theme = plot_theme,
  limits = c(-10.5, 0.5),
  guide = TRUE)

# Yeo-Johnson
p_yj_left <- .create_lambda_shift_plot(
  data = data[distribution == "left-skewed" & method == "Yeo-Johnson"],
  plot_theme = plot_theme,
  limits = c(-10.5, 0.5))

@@

The result is shown in Figure **INSERT REFERENCE**. For each distribution, transformation parameter $\lambda$ varies with $d$ when estimated using the original definitions. In contrast, estimations of $\lambda$ modified power transformations were invariant to $d$.

@@

>>>>

```{r reduced-normality, echo=FALSE, fig.cap=cap}
cap <- paste0(
  "Power transformations may reduce normality depending on location. ",
  "Variable $x$ consists of $10000$ values are randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$. ",
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly dependent on the location $\\mu$, ",
  "and even breaks down due to numerical issues for $\\mu > 5 10^4$.")

@@

x <- power.transform::ragn(10000L, location=0, scale=1/sqrt(2), alpha=0.5, beta=2)
shift_range <- 10^seq(from=0, to=6, by=0.1)

@@

p_bc <- ggplot2::ggplot(
  data = data[method == "Box-Cox"],
  mapping = ggplot2::aes(
    x = mu,
    y = lambda,
    colour = method))
p_bc <- p_bc + plot_theme
p_bc <- p_bc + ggplot2::geom_point()
p_bc <- p_bc + ggplot2::geom_hline(
  yintercept = 1.0, 
  linetype="longdash", 
  colour="grey40")
p_bc <- p_bc + ggplot2::annotate(
  geom = "text",
  x = 6,
  y = 1,
  label = "expected",
  colour=annotation_settings$colour,
  family=annotation_settings$family,
  fontface=annotation_settings$face,
  size=annotation_settings$geom_text_size,
  vjust=-1.0,
  hjust=1.0)
p_bc <- p_bc + ggplot2::scale_x_continuous(
  name = latex2exp::TeX("$\\mu$"),
  labels = scales::math_format())
p_bc <- p_bc + ggplot2::scale_y_continuous(
  name = latex2exp::TeX("$\\lambda$"),
  limits = c(0.0, 35.0))
p_bc <- p_bc + paletteer::scale_color_paletteer_d(
  palette="ggthemes::Tableau_10",
  drop=FALSE,
  guide="none")

p_yj <- ggplot2::ggplot(
  data = data[method == "Yeo-Johnson"],
  mapping = ggplot2::aes(
    x = mu,
    y = lambda,
    colour = method))
p_yj <- p_yj + plot_theme
p_yj <- p_yj + ggplot2::geom_point()
p_yj <- p_yj + ggplot2::geom_hline(
  yintercept = 1.0,
  linetype="longdash",
  colour="grey40")
p_yj <- p_yj + ggplot2::annotate(
  geom = "text",
  x = 6,
  y = 1,
  label = "expected",
  colour=annotation_settings$colour,
  family=annotation_settings$family,
  fontface=annotation_settings$face,
  size=annotation_settings$geom_text_size,
  vjust=-1.0,
  hjust=1.0)
p_yj <- p_yj + ggplot2::scale_x_continuous(
  name = latex2exp::TeX("$\\mu$"),
  labels = scales::math_format())
p_yj <- p_yj + ggplot2::scale_y_continuous(
  name = latex2exp::TeX("$\\lambda$"),
  limits = c(0.0, 35.0))
p_yj <- p_yj + paletteer::scale_color_paletteer_d(
  palette="ggthemes::Tableau_10",
  drop=FALSE)
p_yj <- p_yj + ggplot2::theme(
  axis.title.y=ggplot2::element_blank(),
  axis.text.y=ggplot2::element_blank())

@@

## Modified power transformations are invariant to location

@@

```{r shifted-distributions, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=cap, eval = FALSE}
cap <- paste0(
  "Power transformations modified to be shift-sensitive produce transformation parameters are invariant to location. ")

@@

        for(method in c("box_cox", "yeo_johnson")){
          for(shift in c(FALSE, TRUE)){
            for(robust in c(FALSE)){
            # for(robust in c(FALSE, TRUE)){
              # Only 
              # Skip if robust, but not shifted.
              if(robust && !shift) next

@@

# Update distribution, method and version to factors.
data$distribution <- factor(
  x = data$distribution,
  levels = c("normal", "right-skewed", "left-skewed"))
data$method <- factor(
  x = data$method,
  levels = c("box_cox", "yeo_johnson"),
  labels = c("Box-Cox", "Yeo-Johnson"))
data$version <- factor(
  x = data$version,
  levels = c("original", "shift-sensitive"))

@@

# Box-Cox
p_bc_normal <- .create_lambda_shift_plot(
  data = data[distribution == "normal" & method == "Box-Cox"],
  plot_theme = plot_theme,
  limits = c(0.0, 35.0))

# Yeo-Johnson
p_yj_normal <- .create_lambda_shift_plot(
  data=data[distribution == "normal" & method == "Yeo-Johnson"],
  plot_theme = plot_theme,
  limits = c(0.0, 35.0))

@@

# Box-Cox
p_bc_right <- .create_lambda_shift_plot(
  data =data[distribution == "right-skewed" & method == "Box-Cox"],
  plot_theme = plot_theme,
  limits = c(-9.0, 1.0))

# Yeo-Johnson
p_yj_right <- .create_lambda_shift_plot(
  data = data[distribution == "right-skewed" & method == "Yeo-Johnson"],
  plot_theme = plot_theme,
  limits = c(-9.0, 1.0))

@@

# Box-Cox
p_bc_left <- .create_lambda_shift_plot(
  data = data[distribution == "left-skewed" & method == "Box-Cox"],
  plot_theme = plot_theme,
  limits = c(0.0, 60.0),
  guide = TRUE)

# Yeo-Johnson
p_yj_left <- .create_lambda_shift_plot(
  data = data[distribution == "left-skewed" & method == "Yeo-Johnson"],
  plot_theme = plot_theme,
  limits = c(0.0, 60.0))

@@

The result is shown in Figure **INSERT REFERENCE**, and summarised in Table **INSERT REFERENCE**. For each distribution, transformation parameter $\lambda$ varies with $d$ when estimated using the original definitions. In contrast, estimations of $\lambda$ modified power transformations were invariant to $d$.

@@

## Weighting based on feature values offers best robustness

Outliers may be presented in the data, and affect estimation of power transformation parameters. Such outliers should be down-weighted during the optimisation process. Such weights can be assigned using three different data representations: 1. the original feature values; 2. the feature values after power transformation; 3. the residual errors between transformed feature values and their expected values according to the normal distribution.
